---
title: "Categorical variables and factors"
date: "09-01-2021"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals  
*At the end of this exercise, you will be able to:*    
1. Use the forcats package to manipulate factors, enabling a range of ways to handle categorical variables in your datasets. 

## Questions
If you have question or need help with a specific problem, please [email me](mailto: rerdmann@r.umn.edu).  

## Intro 
In prior sessions, we have learned how to use the `dplyr` package to manipulate dataframes in a number of useful ways - filtering, selecting, and more.  We've also learned how to graph and display data in a number of ways with `ggplot2`.  To add to these abilities, it would be useful to be able to work seamlessly with categorical data - where the values are text-based categories, rather than numbers. To do this, we'll take a closer look at the `forcats` package.



## Install the `forcats` package
```{r message=FALSE, warning=FALSE}
#install.packages("forcats")
```

## Load the `tidyverse`, `forcats`, and `ggplot2`
```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("forcats")
library("ggplot2")
```

## Load the data
For these exercises we will continue using the data from the previous meeting - the "workshop_wide.csv" file.
```{r}
edu_dat <- read_csv("workshop_wide.csv")
```

Let's remind ourselves what the data look like - what variables appear to be categorical?
```{r}
#
```

## Why convert from character to factor?

First, let's look at the frequency of counts within the "likert1_pre" variable.
```{r}
#
```

What things do you notice about this graph and table?  What might we want to change about it?



In order to sequence things using the natural ordering or meaningful sequencing of a categorical variable, we can define the variable as a factor with levels.

To do this for likert1_pre, first we can define the levels in a logical order.

```{r}
#
```

Now, let's create a factor using these levels.

```{r}
#
```

Now, let's try looking at this variable again...

```{r}
#
```

How does this table and plot differ from the first set we made?  Are these changes helpful?  Is there anything more we might want to be able to do?






Let's try to display our "missing" level on our graph.


```{r}
#
```


## Factors within data tables
For this first example, we created a factor from a variable in our edu_dat data table, but we didn't change the likert1_pre variable to be a factor in the data table itself.  What if we want to convert our data table into one containing factors in place of character variables?

```{r}
#
```

Okay, so it looks like this works for our variable in question!  Let's try plotting things again to confirm:

```{r}
#
```

Oh no!  We thought we converted to a factor - what happened?  We didn't define the level ordering, and R will default to alphabetical if you don't tell it otherwise.

Let's try to set the levels to fix our data table... One way to do this is manually.

```{r}
#
```

```{r}
#
```

What if we want to reorder the levels, so that the most frequent response comes first?

```{r}
#
```

What if we wanted to combine factor levels, so that we had three levels for our likert1_pre factor - one for general agreement, one for neutral, and one for general disagreement?

```{r}
#
```

As you can see, fct_recode is useful for both collapsing levels into groups, as well as for renaming factor levels on the fly.


## Practice in breakout rooms!

1. Take a look at edu_dat again. Identify at least 3 variables you could meaningfully convert to factors.  Once you've decided what to convert, make it happen within the data table!  How can you check to ensure the conversion happened?

```{r}
#
```

2. Now that you've converted characters into factors, choose one of these factors and reorder the levels in a way that makes sense to you.  Then, plot the frequency of the factor levels in question.


```{r}
#
```

```{r}
#
```

3. Imagine you enacted a major curriculum redesign in 2018, and you'd like to compare final exam scores before the redesign with scores after the redesign. Use factors to accomplish this grouping and comparison - there's more than one way to successfully display your results!

```{r}

#
```



## Additional resources

Factors chapter in "R for Data Science" - https://r4ds.had.co.nz/factors.html

Official forcats "cheatsheet" - https://raw.githubusercontent.com/rstudio/cheatsheets/master/factors.pdf

That's it for now - great work!